---
title: "CN-PUP SNV Report"
output:
  html_document:
    toc: true
    toc_depth: 3
    number_sections: true
    theme: flatly
params:
  patient_id: "patient"
  long_table: ""
  binary_matrix: ""
  altread_matrix: ""
  refread_matrix: ""
  summary_table: ""
  flowcharts_dir: ""
  breadth_summary: ""
  shared_variant_matrix: ""
  out_prefix: "snv_report"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE, fig.width = 11, fig.height = 7)
suppressPackageStartupMessages({
  library(readr)
  library(dplyr)
  library(tidyr)
  library(stringr)
  library(ggplot2)
  library(DT)
  library(UpSetR)
  library(maftools)
  library(ComplexHeatmap)
  library(circlize)
  library(htmltools)
})

# DT::datatable() embeds the full table as client-side JSON regardless of
# pageLength, which can exceed R's hard 2^31-1 byte string limit at WGS
# scale now that LOWCONF calls are no longer dropped upstream (crashed on
# the "shared by >=2 CTCs" table for PAT-2026-03-00003). Above max_rows,
# skip the interactive widget and point at the TSV/workbook instead --
# stopgap for the current WGS rerun, not a long-term fix for the
# underlying row-count blowup.
safe_datatable <- function(df, ..., max_rows = 50000) {
  if (nrow(df) > max_rows) {
    return(htmltools::tags$blockquote(sprintf(
      "Table skipped: %s rows exceeds the %s-row display cap for this report (interactive tables above this size can exceed R's internal string-size limit). See the long-table TSV / detail workbook for the full data.",
      format(nrow(df), big.mark = ","), format(max_rows, big.mark = ",")
    )))
  }
  datatable(df, ...)
}

patient_id <- params$patient_id
long <- read_tsv(params$long_table, show_col_types = FALSE)
binary <- read_tsv(params$binary_matrix, show_col_types = FALSE)
altread <- read_tsv(params$altread_matrix, show_col_types = FALSE)
refread <- read_tsv(params$refread_matrix, show_col_types = FALSE)
summary_tbl <- read_tsv(params$summary_table, show_col_types = FALSE)

if (!"gene_symbol" %in% names(long)) long$gene_symbol <- ""
if (!"consequence" %in% names(long)) long$consequence <- ""
if (!"impact" %in% names(long)) long$impact <- ""
if (!"QUAL" %in% names(long)) long$QUAL <- ""
for (optional_col in c("hgvsc", "hgvsp", "existing_variation", "max_af", "max_af_pops")) {
  if (!optional_col %in% names(long)) long[[optional_col]] <- ""
}

long <- long %>%
  mutate(
    POS = as.integer(POS),
    DP_num = suppressWarnings(as.numeric(DP)),
    altread_num = suppressWarnings(as.numeric(altread)),
    refread_num = suppressWarnings(as.numeric(refread)),
    # MonoVar QUAL = -log10(P(no variant present anywhere in the cohort at
    # this site)) -- site-level (identical across every cell at a given
    # position), NOT true Phred scale (no x10 factor). Meaningful
    # discriminating range is roughly 0-15.
    QUAL_num = suppressWarnings(as.numeric(QUAL)),
    VAF = if_else((altread_num + refread_num) > 0, altread_num / (altread_num + refread_num), NA_real_),
    gene_display = if_else(is.na(gene_symbol) | gene_symbol == "", paste0("intergenic_", CHROM), gene_symbol),
    Variant_Classification = case_when(
      str_detect(consequence, "missense_variant") ~ "Missense_Mutation",
      str_detect(consequence, "stop_gained") ~ "Nonsense_Mutation",
      str_detect(consequence, "frameshift") ~ "Frame_Shift",
      str_detect(consequence, "splice") ~ "Splice_Site",
      str_detect(consequence, "synonymous_variant") ~ "Silent",
      str_detect(consequence, "intron") ~ "Intron",
      str_detect(consequence, "UTR") ~ "UTR",
      TRUE ~ "Variant"
    ),
    Variant_Type = case_when(
      nchar(REF) == 1 & nchar(ALT) == 1 ~ "SNP",
      nchar(REF) < nchar(ALT) ~ "INS",
      nchar(REF) > nchar(ALT) ~ "DEL",
      TRUE ~ "ONP"
    )
  )

make_maf <- function(df) {
  df %>%
    transmute(
      Hugo_Symbol = gene_display,
      Entrez_Gene_Id = 0,
      Center = "CN-PUP",
      NCBI_Build = "GRCh38",
      Chromosome = str_replace(CHROM, "^chr", ""),
      Start_Position = POS,
      End_Position = POS + pmax(nchar(REF), 1) - 1,
      Strand = "+",
      Variant_Classification = Variant_Classification,
      Variant_Type = Variant_Type,
      Reference_Allele = REF,
      Tumor_Seq_Allele1 = REF,
      Tumor_Seq_Allele2 = ALT,
      Tumor_Sample_Barcode = ctc_id,
      t_ref_count = refread_num,
      t_alt_count = altread_num,
      t_depth = DP_num,
      i_TumorVAF_WU = VAF,
      HGVSp_Short = hgvsp
    )
}

maf_df <- make_maf(long)
maf_path <- paste0(params$out_prefix, ".maf")
write_tsv(maf_df, maf_path)
maf_obj <- tryCatch(read.maf(maf = maf_path, verbose = FALSE), error = function(e) NULL)

mat <- binary %>% as.data.frame()
rownames(mat) <- mat$var_id
mat$var_id <- NULL
mat <- as.matrix(mat)
mode(mat) <- "numeric"
cell_ids <- colnames(mat)
variant_shared_n <- rowSums(mat > 0)
private_variants <- names(variant_shared_n[variant_shared_n == 1])
shared2_variants <- names(variant_shared_n[variant_shared_n >= 2])
all_shared_variants <- names(variant_shared_n[variant_shared_n == length(cell_ids)])

# breadth_summary is an optional reports/<patient>.breadth_summary.tsv from
# SUMMARIZE_BREADTH_QC (only produced when run_bam_qc=true); an empty string
# means it wasn't computed for this run.
breadth_summary_path <- params$breadth_summary
has_breadth <- !is.null(breadth_summary_path) && nzchar(breadth_summary_path) && file.exists(breadth_summary_path)
top_ann <- NULL
if (has_breadth) {
  breadth_raw <- read_tsv(breadth_summary_path, show_col_types = FALSE)
  breadth_cols <- names(breadth_raw)[str_detect(names(breadth_raw), "^breadth_")]
  breadth_col <- c(breadth_cols[str_detect(breadth_cols, "20")], breadth_cols)[1]
  if (!is.na(breadth_col) && "cell_id" %in% names(breadth_raw)) {
    breadth_lookup <- setNames(breadth_raw[[breadth_col]], breadth_raw$cell_id)
    # mat's columns are "<cell_id>__<caller>"; breadth is keyed by raw cell_id.
    cell_base <- str_replace(cell_ids, "__([^_]+)$", "")
    breadth_by_cell <- setNames(as.numeric(breadth_lookup[cell_base]), cell_ids)
    if (any(!is.na(breadth_by_cell))) {
      breadth_by_cell[is.na(breadth_by_cell)] <- 0
      top_ann <- HeatmapAnnotation(
        breadth_at_20x = anno_barplot(breadth_by_cell[cell_ids], ylim = c(0, 1), gp = gpar(fill = "#3567a8")),
        annotation_name_side = "left"
      )
    }
  }
}

flowcharts_dir <- params$flowcharts_dir
flow_summary_path <- file.path(flowcharts_dir, "variant_flow_summary.tsv")
flow_summary <- if (nzchar(flowcharts_dir) && file.exists(flow_summary_path)) {
  read_tsv(flow_summary_path, show_col_types = FALSE)
} else {
  tibble()
}
```

# Overview

```{r overview-table}
overview <- tibble(
  metric = c("patient_id", "cells", "retained_variant_rows", "unique_variants", "variants_shared_by_2_or_more_ctcs", "variants_shared_by_all_ctcs", "private_variants"),
  value = c(patient_id, length(cell_ids), nrow(long), nrow(mat), length(shared2_variants), length(all_shared_variants), length(private_variants))
)
datatable(overview, rownames = FALSE, options = list(dom = 't'))
```

```{r burden}
summary_tbl %>%
  ggplot(aes(x = reorder(ctc_id, variant_count), y = variant_count)) +
  geom_col(fill = "#3567a8") +
  coord_flip() +
  labs(x = "CTC", y = "Retained SNVs", title = "Final SNV burden per CTC") +
  theme_minimal(base_size = 12)
```

# Shared And Private SNVs

```{r upset, fig.height=7}
non_empty_sets <- colnames(mat)[colSums(mat > 0, na.rm = TRUE) > 0]
if (length(non_empty_sets) >= 2 && nrow(mat) > 0 && any(rowSums(mat[, non_empty_sets, drop = FALSE] > 0, na.rm = TRUE) > 0)) {
  upset_df <- as.data.frame((mat[, non_empty_sets, drop = FALSE] > 0) * 1L)
  tryCatch(
    upset(upset_df, nsets = min(length(non_empty_sets), 30), order.by = "freq"),
    error = function(e) cat("UpSet plot skipped: ", conditionMessage(e), sep = "")
  )
} else {
  cat("No shared SNV set structure available for an UpSet plot after filtering.")
}
```

## Interactive Set Table

```{r intersection-table}
if (nrow(mat) > 0 && ncol(mat) > 0) {
  set_table <- tibble(
    var_id = rownames(mat),
    ctc_set = apply(mat > 0, 1, function(x) paste(names(x)[x], collapse = ", ")),
    ctc_count = rowSums(mat > 0)
  ) %>%
    filter(ctc_count > 0) %>%
    group_by(ctc_count, ctc_set) %>%
    summarise(variant_count = n(), variants = paste(var_id, collapse = "; "), .groups = "drop") %>%
    arrange(desc(ctc_count), desc(variant_count), ctc_set)
  datatable(set_table, rownames = FALSE, filter = "top", options = list(pageLength = 20, scrollX = TRUE))
}
```

```{r heatmap, fig.height=9}
if (nrow(mat) > 1 && ncol(mat) > 1) {
  top_vars <- names(sort(variant_shared_n, decreasing = TRUE))[seq_len(min(200, length(variant_shared_n)))]
  Heatmap(mat[top_vars, , drop = FALSE],
          name = "SNV",
          col = c("0" = "white", "1" = "#203864"),
          cluster_rows = TRUE,
          cluster_columns = TRUE,
          show_row_names = FALSE,
          top_annotation = top_ann,
          column_title = "Clustered binary SNV matrix")
}
```

## Pairwise Venn-Style Overlaps

```{r jaccard}
if (ncol(mat) > 1) {
  jac <- matrix(0, ncol(mat), ncol(mat), dimnames = list(cell_ids, cell_ids))
  for (i in seq_along(cell_ids)) {
    for (j in seq_along(cell_ids)) {
      a <- mat[, i] > 0
      b <- mat[, j] > 0
      jac[i, j] <- ifelse(sum(a | b) > 0, sum(a & b) / sum(a | b), 0)
    }
  }
  Heatmap(jac,
          name = "Jaccard",
          col = colorRamp2(c(0, 0.5, 1), c("#f7fbff", "#6baed6", "#08306b")),
          cluster_rows = TRUE,
          cluster_columns = TRUE,
          top_annotation = top_ann,
          column_title = "CTC similarity based on shared SNVs")
}
```

## Shared-Variant Status Matrix

```{r shared-variant-matrix, fig.height=9}
shared_matrix_path <- params$shared_variant_matrix
has_shared_matrix <- !is.null(shared_matrix_path) && nzchar(shared_matrix_path) && file.exists(shared_matrix_path)
if (has_shared_matrix) {
  svm_raw <- read_tsv(shared_matrix_path, show_col_types = FALSE)
}

if (has_shared_matrix && nrow(svm_raw) > 0) {
  svm_raw <- svm_raw %>% arrange(desc(shared_n))
  state_cols <- names(svm_raw)[str_detect(names(svm_raw), "_state$")]
  ctc_names <- str_remove(state_cols, "_state$")

  state_mat <- as.matrix(svm_raw[, state_cols, drop = FALSE])
  colnames(state_mat) <- ctc_names
  rownames(state_mat) <- paste0(
    if_else(is.na(svm_raw$gene_symbol) | svm_raw$gene_symbol == "", "intergenic", svm_raw$gene_symbol),
    " ", svm_raw$CHROM, ":", svm_raw$POS
  )

  # Colors mirror bin/build_detail_workbook.py's CTC_STATE_STYLE/WBC_TIER_STYLE
  # so the report and the companion Excel workbook read consistently.
  state_colors <- c(
    present_pass     = "#0ca30c",
    present_lowconf  = "#c98500",
    filtered_out     = "#6b3fa0",
    absent           = "#d9d9d9",
    not_covered      = "#f2f2f2"
  )
  state_labels <- c(
    present_pass     = "Present (PASS)",
    present_lowconf  = "Present (LOWCONF)",
    filtered_out     = "Removed (germline/pop-AF)",
    absent           = "Absent (0/0)",
    not_covered      = "Not covered"
  )

  wbc_tier_colors <- c(
    clean          = "#0ca30c",
    not_covered    = "#f2f2f2",
    weak_signal    = "#c98500",
    strong_signal  = "#d03b3b",
    not_available  = "#f2f2f2"
  )
  wbc_tier_labels <- c(
    clean          = "WBC clean",
    not_covered    = "WBC not covered",
    weak_signal    = "WBC weak signal",
    strong_signal  = "WBC STRONG signal",
    not_available  = "WBC not available"
  )

  wbc_ann <- rowAnnotation(
    WBC = svm_raw$WBC_tier,
    col = list(WBC = wbc_tier_colors),
    annotation_legend_param = list(WBC = list(at = names(wbc_tier_colors), labels = unname(wbc_tier_labels[names(wbc_tier_colors)])))
  )

  Heatmap(state_mat,
          name = "CTC status",
          col = state_colors,
          cluster_rows = FALSE,
          cluster_columns = FALSE,
          show_row_names = nrow(state_mat) <= 150,
          right_annotation = wbc_ann,
          heatmap_legend_param = list(at = names(state_colors), labels = unname(state_labels[names(state_colors)])),
          column_title = "Shared-variant status matrix (CTC state + WBC germline evidence)")
} else {
  cat("Shared-variant status matrix was not provided to this report.")
}
```

Rows are every variant shared by 2+ CTCs, sorted by how many CTCs share
it. The WBC column on the right applies the same 4-tier germline
check used throughout this session's analysis: a purple "removed"
cell means that CTC's raw call was excluded by germline subtraction
or the population-AF/COSMIC filter (not the depth/alt-read flag) --
worth a second look together with the WBC tier for that row.

```{r pairwise-overlap-table}
if (ncol(mat) > 1) {
  pairwise_overlap <- expand.grid(cell_a = cell_ids, cell_b = cell_ids, stringsAsFactors = FALSE) %>%
    filter(cell_a < cell_b) %>%
    rowwise() %>%
    mutate(
      shared_variants = sum(mat[, cell_a] > 0 & mat[, cell_b] > 0),
      union_variants = sum(mat[, cell_a] > 0 | mat[, cell_b] > 0),
      jaccard = if_else(union_variants > 0, shared_variants / union_variants, 0)
    ) %>%
    ungroup() %>%
    arrange(desc(shared_variants), desc(jaccard))
  datatable(pairwise_overlap, rownames = FALSE, filter = "top", options = list(pageLength = 20, scrollX = TRUE))
}
```

# Variant Flowcharts

```{r flow-summary}
if (nrow(flow_summary)) {
  datatable(flow_summary, rownames = FALSE, filter = "top", options = list(pageLength = 20, scrollX = TRUE))
} else {
  cat("Variant flowchart summary was not provided to this report.")
}
```

```{r mermaid-flowcharts, results='asis'}
if (nrow(flow_summary) && "mermaid_path" %in% names(flow_summary)) {
  cat('<script src="https://cdn.jsdelivr.net/npm/mermaid@10/dist/mermaid.min.js"></script>\n')
  cat('<script>mermaid.initialize({startOnLoad:true, securityLevel:"loose"});</script>\n')
  for (i in seq_len(nrow(flow_summary))) {
    cell <- flow_summary$sample_id[i]
    caller <- flow_summary$caller[i]
    rel <- flow_summary$mermaid_path[i]
    mmd_path <- file.path(flowcharts_dir, rel)
    if (!file.exists(mmd_path)) next
    diagram <- paste(readLines(mmd_path, warn = FALSE), collapse = "\n")
    cat("\n\n## ", htmlEscape(cell), " / ", htmlEscape(caller), "\n\n", sep = "")
    cat('<pre class="mermaid">', htmlEscape(diagram), '</pre>\n', sep = "")
  }
}
```

# Cancer-Style SNV Plots From maftools

```{r maftools-summary}
if (!is.null(maf_obj) && nrow(maf_df) > 0) {
  plotmafSummary(maf = maf_obj, rmOutlier = FALSE, addStat = 'median', dashboard = TRUE)
}
```

```{r oncoplot}
if (!is.null(maf_obj) && nrow(maf_df) > 0) {
  oncoplot(maf = maf_obj, top = min(25, length(unique(maf_df$Hugo_Symbol))), removeNonMutated = TRUE)
}
```

```{r titv}
if (!is.null(maf_obj) && nrow(maf_df) > 0) {
  titv_res <- tryCatch(titv(maf = maf_obj, plot = FALSE, useSyn = TRUE), error = function(e) NULL)
  if (!is.null(titv_res)) plotTiTv(res = titv_res)
}
```

```{r vaf}
if (nrow(long) > 0) {
  long %>%
    filter(!is.na(VAF)) %>%
    ggplot(aes(x = VAF, fill = ctc_id)) +
    geom_histogram(bins = 30, alpha = 0.75, position = "identity") +
    facet_wrap(~ctc_id, scales = "free_y") +
    labs(x = "Variant allele fraction", y = "SNVs", title = "VAF distribution per CTC") +
    theme_minimal(base_size = 11) +
    theme(legend.position = "none")
}
```

```{r dp-hist}
if (nrow(long) > 0) {
  long %>%
    filter(!is.na(DP_num)) %>%
    ggplot(aes(x = DP_num, fill = ctc_id)) +
    geom_histogram(bins = 30, alpha = 0.75, position = "identity") +
    facet_wrap(~ctc_id, scales = "free_y") +
    labs(x = "Depth (DP)", y = "SNVs", title = "Read depth distribution per CTC") +
    theme_minimal(base_size = 11) +
    theme(legend.position = "none")
}
```

```{r altread-hist}
if (nrow(long) > 0) {
  long %>%
    filter(!is.na(altread_num)) %>%
    ggplot(aes(x = altread_num, fill = ctc_id)) +
    geom_histogram(bins = 30, alpha = 0.75, position = "identity") +
    facet_wrap(~ctc_id, scales = "free_y") +
    labs(x = "Alt reads", y = "SNVs", title = "Alt-read count distribution per CTC") +
    theme_minimal(base_size = 11) +
    theme(legend.position = "none")
}
```

```{r qual-hist}
# QUAL is a site-level quantity (identical across every cell at a given
# position), unlike DP/alt-reads which are genuinely per-cell -- de-duplicate
# to one row per unique site first so sites shared across many CTCs aren't
# overweighted just by cell count, and facet by CTC doesn't make sense here
# (every CTC sharing a site would show the exact same bar).
if (nrow(long) > 0) {
  long %>%
    filter(!is.na(QUAL_num)) %>%
    distinct(CHROM, POS, REF, ALT, .keep_all = TRUE) %>%
    ggplot(aes(x = QUAL_num)) +
    geom_histogram(bins = 30, fill = "#2563eb", alpha = 0.85) +
    labs(
      x = "MonoVar QUAL (-log10 P[no variant]; not Phred-scaled)", y = "Unique sites",
      title = "QUAL distribution across unique variant sites"
    ) +
    theme_minimal(base_size = 11)
}
```

# Circos-Like Genome View

```{r circos, fig.height=8, fig.width=8}
if (nrow(long) > 0) {
  circos.clear()
  circos.par(start.degree = 90, gap.degree = 1, track.margin = c(0.01, 0.01), cell.padding = c(0.02, 0, 0.02, 0))
  chrom_df <- long %>%
    mutate(chr_clean = str_replace(CHROM, "^chr", "")) %>%
    group_by(chr_clean) %>%
    summarise(max_pos = max(POS, na.rm = TRUE), .groups = "drop") %>%
    arrange(factor(chr_clean, levels = c(as.character(1:22), "X", "Y", "M", "MT")))
  circos.initialize(factors = chrom_df$chr_clean, xlim = cbind(rep(0, nrow(chrom_df)), chrom_df$max_pos))
  circos.trackPlotRegion(ylim = c(0, 1), panel.fun = function(x, y) {
    chr <- CELL_META$sector.index
    circos.text(CELL_META$xcenter, CELL_META$ylim[2], chr, cex = 0.6, facing = "clockwise", niceFacing = TRUE)
  }, bg.border = NA, track.height = 0.06)
  plot_df <- long %>% mutate(chr_clean = str_replace(CHROM, "^chr", ""), shared_n = variant_shared_n[var_id])
  colors <- ifelse(plot_df$shared_n == length(cell_ids), "#b2182b", ifelse(plot_df$shared_n >= 2, "#2166ac", "#999999"))
  circos.trackPlotRegion(ylim = c(0, max(plot_df$shared_n, na.rm = TRUE) + 1), panel.fun = function(x, y) {
    chr <- CELL_META$sector.index
    dd <- plot_df[plot_df$chr_clean == chr, ]
    if (nrow(dd) > 0) circos.points(dd$POS, dd$shared_n, col = colors[plot_df$chr_clean == chr], pch = 16, cex = 0.45)
  }, bg.border = "#dddddd", track.height = 0.22)
  legend("bottomleft", legend = c("all CTCs", ">=2 CTCs", "private"), col = c("#b2182b", "#2166ac", "#999999"), pch = 16, bty = "n", cex = 0.8)
  circos.clear()
}
```

# Shared Variant Tables

## Variants Shared By All CTCs

```{r all-shared-table}
all_shared_table <- long %>%
  filter(var_id %in% all_shared_variants) %>%
  arrange(CHROM, POS, ctc_id) %>%
  select(any_of(c("ctc_id", "var_id", "CHROM", "POS", "REF", "ALT", "gene_symbol", "consequence", "impact", "hgvsp", "existing_variation", "max_af", "max_af_pops", "DP", "altread", "VAF")))
safe_datatable(all_shared_table, rownames = FALSE, filter = "top", options = list(pageLength = 20, scrollX = TRUE))
```

## Variants Shared By At Least Two CTCs

```{r shared2-table}
shared2_table <- long %>%
  filter(var_id %in% shared2_variants) %>%
  mutate(shared_ctc_count = variant_shared_n[var_id]) %>%
  arrange(desc(shared_ctc_count), CHROM, POS, ctc_id) %>%
  select(any_of(c("shared_ctc_count", "ctc_id", "var_id", "CHROM", "POS", "REF", "ALT", "gene_symbol", "consequence", "impact", "hgvsp", "existing_variation", "max_af", "max_af_pops", "DP", "altread", "VAF")))
safe_datatable(shared2_table, rownames = FALSE, filter = "top", options = list(pageLength = 20, scrollX = TRUE))
```

Need the exact per-cell DP/alt-reads behind any variant above (including the WBC's own reads)? See the companion workbook: **`r patient_id`_detail.xlsx**, generated alongside this report with one row per variant/cell plus a WBC lookup for every shared variant.

# Per-CTC Pages

```{r per-ctc, results='asis'}
for (cell in cell_ids) {
  cat("\n\n## ", cell, "\n\n", sep = "")
  dd <- long %>% filter(ctc_id == cell) %>% arrange(CHROM, POS)
  cat("Retained SNVs: ", nrow(dd), "\n\n", sep = "")
  if (nrow(dd) > 0) {
    print(
      dd %>%
        select(any_of(c("var_id", "CHROM", "POS", "REF", "ALT", "gene_symbol", "consequence", "impact", "hgvsp", "existing_variation", "max_af", "max_af_pops", "DP", "refread", "altread", "VAF", "GT", "FILTER"))) %>%
        safe_datatable(rownames = FALSE, filter = "top", options = list(pageLength = 15, scrollX = TRUE))
    )
  }
}
```
